Quantifying the importance of an inference model in Bayesian phylogenetics

### 2020-06-08 pirouette_supplement.tex
Changes @Giappo

\begin{itemize}
  \item{
    subsection \ref{subsec:guidelines}: guidelines for users
  }
  \item{
    subsection \ref{subsec:installation}: installation
  }
  \item{
    subsection \ref{subsec:resources}: resources, such as 
    website, tutorials, packages used, bug reporting and contributing
  }
  \item{
    subsection \ref{subsec:citation}: citation of pirouette
  }
  \item{
    subsection \ref{subsec:main_functions}: main functions
  }
  \item{
    subsection \ref{subsec:main_example}: code, extra figures and
    diagnostics regarding the main example.
  }
  \item{
    subsection \ref{subsec:distribution}: the result of using 
    multiple trees, as generated by the
    same stochastic process as the main example
  }
  \item{
    subsection \ref{subsec:n_taxa}: the effect of the number
    of taxa
  }
  \item{
    subsection \ref{subsec:n_nucleotides}: the effect of the DNA
    alignment sequence length
  }
  \item{
    subsection \ref{subsec:simplest_correct_parameterization} shows the
    effect when doing an inference in the simplest use case
  }
  \item{
    subsection \ref{subsec:under_parameterization} shows the
    effect when doing an inference with an under-parameterization
  }
  \item{
    subsection \ref{subsec:different_n_mutations} shows the
    effect when the twin alignment is allowed to have a different
    number of mutations
  }
  \item{
    subsection \ref{subsec:mutation_rate} shows the
    effect of different mutation rates
  }
\end{itemize}

@@

\begin{itemize}
  \item{
    The use of 20 replicates results in decent plots.
  }
  \item{
    The use of more taxa increases the inference error
  }
  \item{
    The use of longer DNA sequences decreases the inference error.
  }
  \item{
    When we do not impose the same number of mutations between true and twin alignment, we observe a difference in the error distributions with respect to the standard case (presented in the main text) where they are forced to have the same number of mutations.
  }
  \item{
    Using a mutation rate less than 1.0 / crown age, decreases the
    inference error. We predict this will increase the error in the parameter
    estimation.
  }
\end{itemize}

@@

@@

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_28/errors.png}
  \caption{
    Aggregate error distributions, similar to Fig. \ref{fig:example_30} 
    for the main example, but now for a collection of 20 replicate trees. 
    For each setting (true generative, true best candidate, twin generative 
    and twin best candidate), the resulting errors from each replicate 
    pipeline have been merged into a single distribution.
  }
  \label{fig:replicate_trees}
\end{figure}

@@

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_28/errors.png}
  \caption{Aggregate error distributions for 20 replicates. Here each true tree has 6 taxa. This took 16 hours to compute.}
  \label{fig:example_6_taxa}
\end{figure}

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_32/errors.png}
  \caption{Aggregate error distributions for 10 replicates. Here each true tree has 12 taxa. This took 19 hours to compute.}
  \label{fig:example_12_taxa}
\end{figure}

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_33/errors.png}
    \caption{Aggregate error distributions for 20 replicates. Here each true tree has 24 taxa. This took 28 hours to compute.}
    \label{fig:example_24_taxa}
\end{figure}

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_41/errors.png}
  \caption{Aggregate error distributions for 5 replicates. Here each true tree has 32 taxa. This took 7 hours to compute.}
  \label{fig:example_32_taxa}
\end{figure}

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_42/errors.png}
  \caption{Aggregate error distributions for 2 replicates. Here each true tree has 40 taxa. This took 8 hours to compute.}
  \label{fig:example_40_taxa}
\end{figure}

@@

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_19/errors.png}
  \caption{Aggregate error distributions for 20 replicates. Here each each alignment has a sequence length of 500 nucleotides. This took 12 hours to compute.}
  \label{fig:example_500_nucleotides}
\end{figure}

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_28/errors.png}
  \caption{Aggregate error distributions for 20 replicates. Here each each alignment has a sequence length of 1000 nucleotides (as used in the main example). This took 16 hours to compute.}
  \label{fig:example_1000_nucleotides}
\end{figure}

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_34/errors.png}
  \caption{Aggregate error distributions for 10 replicates. Here each each alignment has a sequence length of 2000 nucleotides. This took 11 hours to compute.}
  \label{fig:example_2000_nucleotides}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Effect of assuming a Yule tree prior on a Yule tree}
\label{subsec:simplest_correct_parameterization}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_22/errors.png}
  \caption{Aggregate error distributions for 10 replicates. Here each true tree is generated by a Yule process. For the inference we used a Yule tree prior. This took 9 hours to compute.}
  \label{fig:example_yule}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Effect of assuming a Yule tree prior on a BD tree}
\label{subsec:under_parameterization}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_26/errors.png}
  \caption{Aggregate error distributions for 20 replicates. Here each true tree is generated by a BD process. For the inference we used instead a Yule tree prior. This took 9 hours to compute.}
\end{figure}

@@

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_23/errors_low.png}
  \caption{Aggregate error distributions for a distribution of trees, where the true trees are DD with low likelihood.}
\end{figure}

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_23/errors_mid.png}
  \caption{Aggregate error distributions for a distribution of trees, where the true trees are DD with median likelihood.}
\end{figure}

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_23/errors_high.png}
  \caption{Aggregate error distributions for a distribution of trees, where the true trees are DD with high likelihood.}
\end{figure}

@@

The main example uses a twin alignment that has the same number
of mutations (as measured from the ancestral sequence) as the true alignment. Here, we show the same results, with the difference that
the twin alignment uses the same mutation rate, yet is not guaranteed
to have the same number of mutations. Any difference in the error distributions can therefore also stem from a difference in the amount of information contained in the alignments.

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_18/errors.png}
  \caption{Aggregate error distributions similar to Fig. \ref{fig:replicate_trees}, but here the number of mutations is not imposed to be the same between true and twin alignment. Instead we just use an equal mutation rate. This took 10 hours to compute.}
  \label{fig:example_random_mutations}
\end{figure}

\new{
Comparing Fig. \ref{fig:example_random_mutations} to  \label{fig:replicate_trees} we can see that the extent of the error tend to increase both for the true and the twin distributions. This is probably due to the fact that 
}

@@

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_35/errors.png}
  \caption{Aggregate error distributions for the tree distribution presented in \ref{subsec:distribution} but with a per-nucleotide mutation rate of 0.25 / crown age. This took 11 hours to compute.}
\end{figure}

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_36/errors.png}
  \caption{Aggregate error distributions for the tree distribution presented in \ref{subsec:distribution} but with a per-nucleotide mutation rate of 0.50 / crown age. This took 14 hours to compute.}
\end{figure}

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_37/errors.png}
  \caption{Aggregate error distributions for the tree distribution presented in \ref{subsec:distribution} but with a per-nucleotide mutation rate of 0.75 / crown age. This took 16 hours to compute.}
\end{figure}

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_38/errors.png}
  \caption{Aggregate error distributions for the tree distribution presented in \ref{subsec:distribution} but with a per-nucleotide mutation rate of 1.25 / crown age. This took 17 hours to compute.}
\end{figure}

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_39/errors.png}
  \caption{Aggregate error distributions for the tree distribution presented in \ref{subsec:distribution} but with a per-nucleotide mutation rate of 1.50 / crown age. This took 10 hours to compute.}
\end{figure}

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_40/errors.png}
  \caption{Aggregate error distributions for the tree distribution presented in \ref{subsec:distribution} but with a per-nucleotide mutation rate of 2.0 / crown age. This took 19 hours to compute.}
\end{figure}

>>>>

\begin{itemize}
  \item{
    subsection \ref{subsec:guidelines}: guidelines for users
  }
  \item{
    subsection \ref{subsec:installation}: installation
  }
  \item{
    subsection \ref{subsec:resources}: resources, such as 
    website, tutorials, packages used, bug reporting and contributing
  }
  \item{
    subsection \ref{subsec:citation}: citation of pirouette
  }
  \item{
    subsection \ref{subsec:main_functions}: main functions
  }
  \item{
    subsection \ref{subsec:main_example}: code, extra figures and
    diagnostics regarding the main example.
  }
  \item{
    subsection \ref{subsec:distribution}: the result of using 
    multiple trees, as generated by the
    same stochastic process as the main example
  }
  \item{
    subsection \ref{subsec:n_taxa}: the effect of the number
    of taxa
  }
  \item{
    subsection \ref{subsec:n_nucleotides}: the effect of the DNA
    alignment sequence length
  }
  \item{
    subsection \ref{subsec:simplest_correct_parameterization} shows the
    effect when doing an inference in the simplest use case
  }
  \item{
    subsection \ref{subsec:under_parameterization} shows the
    effect when doing an inference with an under-parameterization
  }
  \item{
    subsection \ref{subsec:different_n_mutations} shows the
    effect when the twin alignment is allowed to have a different
    number of \new{substitution}s
  }
  \item{
    subsection \ref{subsec:mutation_rate} shows the
    effect of different mutation rates
  }
\end{itemize}

@@

\begin{itemize}
  \item{
    The use of 20 replicates results in decent plots.
  }
  \item{
    The use of more taxa increases the inference error
  }
  \item{
    The use of longer DNA sequences decreases the inference error.
  }
  \item{
    When we do not impose the same number of \new{substitution}s between true and twin alignment, we observe a difference in the error distributions with respect to the standard case (presented in the main text) where they are forced to have the same number of \new{substitution}s.
  }
  \item{
    Using a mutation rate less than 1.0 / crown age, decreases the
    inference error. We predict this will increase the error in the parameter
    estimation.
  }
\end{itemize}

@@

\clearpage
\newpage

@@

\begin{figure}[H]
  \includegraphics[width=0.98\textwidth]{pirouette_example_28/errors.png}
  \caption{
    Aggregate error distributions, similar to Fig. \ref{fig:example_30} 
    for the main example, but now for a collection of 20 replicate trees. 
    For each setting (true generative, true best candidate, twin generative 
    and twin best candidate), the resulting errors from each replicate 
    pipeline have been merged into a single distribution.
  }
  \label{fig:replicate_trees}
\end{figure}

@@

\begin{figure}[H]
  \includegraphics[width=0.98\textwidth]{pirouette_example_28/errors.png}
  \caption{Aggregate error distributions for 20 replicates. Here each true tree has 6 taxa. This took 16 hours to compute.}
  \label{fig:example_6_taxa}
\end{figure}

\begin{figure}[H]
  \includegraphics[width=0.98\textwidth]{pirouette_example_32/errors.png}
  \caption{Aggregate error distributions for 10 replicates. Here each true tree has 12 taxa. This took 19 hours to compute.}
  \label{fig:example_12_taxa}
\end{figure}

\begin{figure}[H]
  \includegraphics[width=0.98\textwidth]{pirouette_example_33/errors.png}
    \caption{Aggregate error distributions for 20 replicates. Here each true tree has 24 taxa. This took 28 hours to compute.}
    \label{fig:example_24_taxa}
\end{figure}

\begin{figure}[H]
  \includegraphics[width=0.98\textwidth]{pirouette_example_41/errors.png}
  \caption{Aggregate error distributions for 5 replicates. Here each true tree has 32 taxa. This took 7 hours to compute.}
  \label{fig:example_32_taxa}
\end{figure}

\begin{figure}[H]
  \includegraphics[width=0.98\textwidth]{pirouette_example_42/errors.png}
  \caption{Aggregate error distributions for 2 replicates. Here each true tree has 40 taxa. This took 8 hours to compute.}
  \label{fig:example_40_taxa}
\end{figure}

@@

\begin{figure}[H]
  \includegraphics[width=0.98\textwidth]{pirouette_example_19/errors.png}
  \caption{Aggregate error distributions for 20 replicates. Here each each alignment has a sequence length of 500 nucleotides. This took 12 hours to compute.}
  \label{fig:example_500_nucleotides}
\end{figure}

\begin{figure}[H]
  \includegraphics[width=0.98\textwidth]{pirouette_example_28/errors.png}
  \caption{Aggregate error distributions for 20 replicates. Here each each alignment has a sequence length of 1000 nucleotides (as used in the main example). This took 16 hours to compute.}
  \label{fig:example_1000_nucleotides}
\end{figure}

\begin{figure}[H]
  \includegraphics[width=0.98\textwidth]{pirouette_example_34/errors.png}
  \caption{Aggregate error distributions for 10 replicates. Here each each alignment has a sequence length of 2000 nucleotides. This took 11 hours to compute.}
  \label{fig:example_2000_nucleotides}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The effect of assuming a Yule tree prior on a Yule tree}
\label{subsec:simplest_correct_parameterization}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{figure}[H]
  \includegraphics[width=0.98\textwidth]{pirouette_example_22/errors.png}
  \caption{Aggregate error distributions for 10 replicates. Here each true tree is generated by a Yule process. For the inference we used a Yule tree prior. This took 9 hours to compute.}
  \label{fig:example_yule}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The effect of assuming a Yule tree prior on a BD tree}
\label{subsec:under_parameterization}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{figure}[H]
  \includegraphics[width=0.98\textwidth]{pirouette_example_26/errors.png}
  \caption{Aggregate error distributions for 20 replicates. Here each true tree is generated by a BD process. For the inference we used instead a Yule tree prior. This took 9 hours to compute.}
\end{figure}

@@

\begin{figure}[H]
  \includegraphics[width=0.98\textwidth]{pirouette_example_23/errors_low.png}
  \caption{Aggregate error distributions for a distribution of trees, where the true trees are DD with low likelihood.}
\end{figure}

\begin{figure}[H]
  \includegraphics[width=0.98\textwidth]{pirouette_example_23/errors_mid.png}
  \caption{Aggregate error distributions for a distribution of trees, where the true trees are DD with median likelihood.}
\end{figure}

\begin{figure}[H]
  \includegraphics[width=0.98\textwidth]{pirouette_example_23/errors_high.png}
  \caption{Aggregate error distributions for a distribution of trees, where the true trees are DD with high likelihood.}
\end{figure}

@@

The main example uses a twin alignment that has the same number
of \new{substitution}s (as measured from the ancestral sequence) as the true alignment. Here, we show the same results, with the difference that
the twin alignment uses the same mutation rate, yet is not guaranteed
to have the same number of \new{substitution}s.
\iffalse
Any difference in the error distributions can therefore also stem from a difference in the amount of information contained in the alignments.
\fi

\begin{figure}[H]
  \includegraphics[width=0.98\textwidth]{pirouette_example_18/errors.png}
  \caption{Aggregate error distributions similar to Fig. \ref{fig:replicate_trees}, but here the number of \new{substitution}s is not imposed to be the same between true and twin alignment. Instead we just use an equal mutation rate. This took 10 hours to compute.}
  \label{fig:example_random_mutations}
\end{figure}

\new{
Comparing figures \ref{fig:example_random_mutations} and \ref{fig:replicate_trees} we can see that the discrepancy between true and twin distributions tend to increase. This is probably due to the fact that letting mutation rates induces a difference in the amount of information contained in the alignments and this is reflected in the error distributions.
}

@@

\begin{figure}[H]
  \includegraphics[width=0.98\textwidth]{pirouette_example_35/errors.png}
  \caption{Aggregate error distributions for the tree distribution presented in \ref{subsec:distribution} but with a per-nucleotide mutation rate of 0.25 / crown age. This took 11 hours to compute.}
\end{figure}

\begin{figure}[H]
  \includegraphics[width=0.98\textwidth]{pirouette_example_36/errors.png}
  \caption{Aggregate error distributions for the tree distribution presented in \ref{subsec:distribution} but with a per-nucleotide mutation rate of 0.50 / crown age. This took 14 hours to compute.}
\end{figure}

\begin{figure}[H]
  \includegraphics[width=0.98\textwidth]{pirouette_example_37/errors.png}
  \caption{Aggregate error distributions for the tree distribution presented in \ref{subsec:distribution} but with a per-nucleotide mutation rate of 0.75 / crown age. This took 16 hours to compute.}
\end{figure}

\begin{figure}[H]
  \includegraphics[width=0.98\textwidth]{pirouette_example_38/errors.png}
  \caption{Aggregate error distributions for the tree distribution presented in \ref{subsec:distribution} but with a per-nucleotide mutation rate of 1.25 / crown age. This took 17 hours to compute.}
\end{figure}

\begin{figure}[H]
  \includegraphics[width=0.98\textwidth]{pirouette_example_39/errors.png}
  \caption{Aggregate error distributions for the tree distribution presented in \ref{subsec:distribution} but with a per-nucleotide mutation rate of 1.50 / crown age. This took 10 hours to compute.}
\end{figure}

\begin{figure}[H]
  \includegraphics[width=0.98\textwidth]{pirouette_example_40/errors.png}
  \caption{Aggregate error distributions for the tree distribution presented in \ref{subsec:distribution} but with a per-nucleotide mutation rate of 2.0 / crown age. This took 19 hours to compute.}
\end{figure}

\giovanni{I am not able to give an interpretation to these figures. I would expect a trend between the difference of true and twin medians and mutation rate, but I cannot see anything. The fact that sometimes true is smaller and some other times twin is, makes me think that these results are probably not very reliable, probably due to a not sufficient number of replicates.}
